---
title: "Open crime data (in R)"
author: "Matt Ashby"
date: "9 December 2020"
output: github_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

These brief notes are for the workshop 'Open crime data in R' at the [Methods@Manchester](https://www.methods.manchester.ac.uk/) hackathon on 'Exploring the Relationship between Crowds, Safety and Crime', held remotely on 9 December 2020.


## What is open crime data?

"[Open data and content can be freely used, modified, and shared by anyone for any purpose](https://opendefinition.org/)", typically under a [Creative Commons](https://creativecommons.org/) or similar licence. Open crime data typically refers to incident-level data about crime (or related topics such as police activity) that is made available under an open-data licence. 


## Why use open crime data?

Open crime data is often much faster to access, compared to obtaining data directly from an agency. Negotiating data-sharing agreement and obtaining security clearances for researchers often delays research for months (or even years), whereas open data can be downloaded immediately it is released (and many agencies update their open data daily or weekly). This means open data can be used to carry out fast-time research, which can be particularly useful where an answer to a research question is needed quickly (e.g. to respond to a government consultation).

Not needing to negotiate data access directly with agencies also means that it is more feasible to do comparative research using data from multiple agencies. This is important because it allows us to identify how generaliable the results of our work are to other settings.

Since open crime data can be accessed by anyone, it does not have to be handled on secure servers or the results restricted to a few people. This can be particularly useful when using data internationally, since there is no need to comply with restrictions on the handling of sensitive data that existing in many countries.


## Why not to use open crime data

If you need data fields that aren't included in the open data provided by a particular agency, obviously open data is of no use to you. It's very rare for agencies to release all the fields in their own data as part of their open data extracts, so there will almost always be more fields available if you approach the agency directly.

Open data is typically not suitable for research that is using an individual or a point location as the unit of analysis, since names are typically redacted ([although not in Dallas](https://www.dallasopendata.com/Public-Safety/Police-Incidents/qv6i-rri7)) and locations often obfuscated (e.g. by snapping locations to the centroid of the nearest street segment) to maintain victim privacy.


## A tour of open crime data sources

For England, Wales and Northern Ireland (but not Scotland), [data.police.uk](https://data.police.uk) provides incident-level open crime data, but there is no date stamp (only the month in which the crime occurred) and offences are grouped into a small number of broad categories.

The [Police Data Initiative](https://www.policedatainitiative.org) provides a directory of open crime and other police data (e.g. arrests, traffic stops) from various US agencies, but there are lots of US agencies that produce open crime data but aren't listed in the directory.

The [Open Data Network](http://www.opendatanetwork.com) provides a tool for searching the metadata of datasets hosted by Socrata, the main provider of open-data platform software in the US, but there is no advanced search feature so this often involves trawling through large numbers of results. Note that many agencies refer to their crime records as 'incidents' rather than 'crimes'.

[ArcGIS Hub](https://hub.arcgis.com/search) provides a similar search tool for open-data websites that they host, although again the tagging of metadata by different agencies is very hit and miss, so it's often better to do a more general search and then browse through the results manually.

Some US agencies provide open data but don't use one of the big platforms (e.g. [Portland, OR](https://www.portlandoregon.gov/police/76454)), making their data easier to find by Googling '[city name] open data' and browsing from there.

The [Crime Open Database](https://osf.io/zyaqn/) provides harmonised crime data for 16 large US cities, although the process of harmonising data from different cities operating under different laws is [inevitably imperfect](https://doi.org/10.1163/24523666-00401007).


## Using the `crimedata` package in R

The [Crime Open Database](https://osf.io/zyaqn/) (CODE) contains about 22.5 million crime records from 16 of the largest cities in the United States from 2007 onwards. Each record includes the type, date/time and approximate location of a crime.

You can access CODE directly in R using the `crimedata` package. Use the function `get_crime_data()` to download data for specific cities or specific years:

```{r}
# install package if not already installed
# install.packages("crimedata")

library("crimedata")

crime_data <- get_crime_data(years = 2019, cities = "San Francisco")
```

The first call to `get_crime_data()` takes a while to run (you may need to get a coffee) because it downloads a list of URLs in the background for the current locations of the data files. This is only done once per session, so subsequent calls are faster.

Some cities do not provide as many data fields as others, so you may see (as above) a warning that some columns (in this case, `location_type` and `location_category`) are missing from the data you have requested – this is not an error with your code, just a reflection of San Francisco open crime data not including a location-type field.

```{r}
crime_data
```

Because the data available is quite large, by default `get_crime_data()` downloads only a 10% sample of the crime data requested. This allows you to refine your request before downloading the full dataset by adding the `type = "core"` parameter to your function.

There is a lot more information about CODE data, including details of the crime and location categories, on the [CODE project website](https://osf.io/zyaqn/wiki).

